A catalog of wide binaries and comoving star pairs from \textsl{Gaia DR1} and low-resolution spectroscopy

### 2017-07-04 paper/comoving-followup.tex
new figure wooooooo
words

% Project-specific
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\DR}[1]{\acronym{DR}#1}
\newcommand{\tgas}{\acronym{TGAS}}
\newcommand{\lbfgsb}{\texttt{L-BFGS-B}}
\newcommand{\Ha}{\ensuremath{{\rm H}\alpha}}
\newcommand{\dex}{\ensuremath{{\rm dex}}}

@@

We observed a total of 765 unique sources over 5 nights from (UT) 2017-03-11 to
2017-03-15, corresponding to 321 unique star pairs and $\approx 120$
calibration targets.
The calibration targets have previously measured radial velocities and chemical
abundances, which will be used for validation of our reduction pipeline and in
future work to measure metallicities and stellar parameters for each observed
source using \project{The Cannon} (\citealt{Ness:TODO}).
The spectra were reduced and calibrated using a custom, publicly-available
pipeline written in \python, described below.

@@

\subsection{Validation}

Several of our targets and all of our calibration targets have previously
measured radial velocities (RVs).
We can therefore estimate the absolute velocity uncertainties by comparing our
derived radial velocities with high-quality prior measurements.
Following previous work (\citealt{Shaya:2011}), we search for previous RV
measurements (in decreasing order of priority) with:
\begin{enumerate}
  \item The Geneva-Copenhagen Survey of the Solar Neighborhood
  (\citealt{Nordstrom:2004}).
  \item The Catalogue of Radial Velocities of Galactic Stars with Astrometric
  Data, the Second Version (\citealt{Kharchenko:2007}).
  \item The General Catalogue of Mean Radial Velocities
  (\citealt{Barbier-Brossat:2000}).
  \item The SIMBAD database (\citealt{Wenger:2000}). We use \package{astroquery}
  to retrieve records from the database, which do not contain uncertainties but
  do contain a quality flag. We only retain RVs that have a quality flag set to
  `A' and adopt a $5~\kms$ uncertainty for all such values.
\end{enumerate}
We find that \todo{XXX} of our target stars have reliable prior measurements,
along with the \todo{YYY} calibration targets.
\figurename~\ref{fig:compare-previous} shows the distribution of velocity
differences between our measurements and literature values.
From this distribution, we compute the MAD and estimate the standard deviation
to be $\approx 20~\kms$, which we adopt as our global systematic error for all
absolute radial velocities.
We note, however, that for computing \emph{relative} velocities for pairs of
nearby stars, it is the RV precision that matters.

% \begin{figure}[htbp]
%   \begin{center}
%     \includegraphics[width=\linewidth]{compare_previous.pdf}
%   \end{center}
%   \caption{%
%     \todo{Plot showing my RV's vs. previously measured.}
%     \label{fig:compare-previous}}
% \end{figure}

@@

\subsection{Catalog ...}\label{sec:}

\todo{10 overlap with Shaya and Olling}

@@

\todo{Plot of magnitude of 3d velocity difference vs. 3d separation (inferred).}

\todo{separation distribution - combine with Shaya and Olling sample!}

>>>>

% Project-specific
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\DR}[1]{\acronym{DR}#1}
\newcommand{\tgas}{\acronym{TGAS}}
\newcommand{\lbfgsb}{\texttt{L-BFGS-B}}
\newcommand{\Ha}{\ensuremath{{\rm H}\alpha}}
\newcommand{\dex}{\ensuremath{{\rm dex}}}
\newcommand{\llrold}{\ensuremath{\mathcal{R}_\mu}}
\newcommand{\llrnew}{\ensuremath{\mathcal{R}_{\rm RV}}}

@@

We observed a total of 765 unique sources over 5 nights from (UT) 2017-03-11 to
2017-03-15, corresponding to 323 unique star pairs and $\approx 120$
calibration targets.
The calibration targets have previously measured radial velocities and chemical
abundances, which will be used for validation of our reduction pipeline and in
future work to measure metallicities and stellar parameters for each observed
source using \project{The Cannon} (\citealt{Ness:TODO}).
The spectra were reduced and calibrated using a custom, publicly-available
pipeline written in \python, described below.

@@

\subsection{Validation of absolute radial velocities}

Several of our targets and all of our calibration targets have previously
measured radial velocities (RVs).
We can therefore estimate the absolute velocity uncertainties by comparing our
derived radial velocities with high-quality prior measurements.
Following previous work (\citealt{Shaya:2011}), we search for previous RV
measurements (in decreasing order of priority) with:
\begin{enumerate}
  \item The Geneva-Copenhagen Survey of the Solar Neighborhood
  (\citealt{Nordstrom:2004}).
  \item The Catalogue of Radial Velocities of Galactic Stars with Astrometric
  Data, the Second Version (\citealt{Kharchenko:2007}).
  \item The General Catalogue of Mean Radial Velocities
  (\citealt{Barbier-Brossat:2000}).
  \item The SIMBAD database (\citealt{Wenger:2000}). We use \package{astroquery}
  to retrieve records from the database, which do not contain uncertainties but
  do contain a quality flag. We only retain RVs that have a quality flag set to
  `A' and adopt a $5~\kms$ uncertainty for all such values.
\end{enumerate}
% See: paper/figures/Compare with previous observations.ipynb
We find that 155 of the observed stars (mainly calibration targets) have
reliable prior measurements from the literature search described above.
\figurename~\ref{fig:compare-previous} shows the distribution of velocity
differences between our measurements and literature values.
From this distribution, we compute the median absolute deviation (MAD) and
estimate the standard deviation to be $\approx 20~\kms$, which we adopt as our
global systematic error for all absolute radial velocities.
We note, however, that for computing \emph{relative} velocities for pairs of
nearby stars, it is the RV precision (typically $\approx 5$--$10~\kms$) that
matters.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width=\linewidth]{rv-comparison.pdf}
  \end{center}
  \caption{%
    \todo{Plot showing my RV's vs. previously measured.}
    \label{fig:compare-previous}}
\end{figure}

@@

\subsection{Comoving pair catalog}\label{sec:catalog}

Of the 323 observed star pairs, 11 have at least one member with an unreliable
(flagged) radial velocity measurement from the pipeline above.
For the remaining 312 pairs, we use the measured radial velocities to re-compute
the marginal likelihood ratio for the two hypotheses considered in previous work
(see Appendix~B, \citealt{Oh:2017}).
We compute two fully marginalized likelihoods (FMLs) for the data: (1) assuming
that the two stars are comoving with identical 3D velocity vectors (within some
allowed small dispersion), and (2) assuming that the two stars have independent
3D velocity vectors drawn from a prior velocity distribution.
As with previous work, we set the small dispersion to the orbital velocity of a
binary with the separation of the pair, and use a mixture of three Gaussian
velocity components for the prior, meant to represent the thin disk, thick disk,
and halo velocity distributions relative to the sun.

To compute the new likelihood ratio, we modify the expressions from
\citet{Oh:2017} to include the radial velocity measurements.
In the same notation, this requires the following changes:
\begin{itemize}
  \item The projection matrix, $\mat{M}$, now has to include the radial unit
    vector
    \begin{equation}
      \mat{M} =
        \left(
          \begin{array}{c c c}
            -\sin\alpha & \cos\alpha & 0 \\
            -\sin\delta \, \cos\alpha & -\sin\delta \, \sin\alpha & \cos\delta\\
            \cos\delta \, \cos\alpha & \cos\delta \, \sin\alpha & \sin\delta
          \end{array}
        \right)
    \end{equation}
    where $(\alpha, \delta)$ are the ICRS coordinates (right ascension and
    declination) of a given star.
  \item The data-space velocity vector, $\vec{y}$, for one star now becomes
    \begin{equation}
      \vec{y} =
        \transpose{\left(
          \begin{array}{c c c}
            r_i\,\mu_{\alpha}^* &
            r_i\,\mu_{\delta} &
            v_{r}
          \end{array}
        \right)}
    \end{equation}
    where $\mu_\alpha^*$ is the proper motion in right ascension (including the
    $\cos{\delta}$ term), $\mu_\delta$ is the proper motion in declination, and
    $v_r$ is the radial velocity.
  \item The covariance matrix becomes the block-diagonal matrix
    \begin{equation}
      \mat{\Sigma} = \left(
        \begin{array}{c c}
          r^2 \, \mat{C} & 0 \\
          0 & \sigma_{v_r}^2
        \end{array}
      \right)
    \end{equation}
    where $r$ is the distance, $\mat{C}$ is the proper motion covariance matrix
    (from the \tgas\ catalog), and $\sigma_{v_r}$ is the radial velocity
    uncertainty.
\end{itemize}
To compute the FML for hypothesis 1 for a pair of stars $(i,j)$, the data-space
vector must become the 6-vector
\begin{equation}
  \vec{y} = \transpose{\left(
    \begin{array}{c c}
      \vec{y}_i &
      \vec{y}_j
    \end{array}
    \right)}
\end{equation}
and the covariance matrix becomes the block-diagonal matrix
\begin{equation}
  \mat{\Sigma} = \left(
    \begin{array}{c c}
      \mat{\Sigma}_i & 0 \\
      0 & \mat{\Sigma}_j
    \end{array}
  \right) \quad .
\end{equation}

If the new likelihood ratio is larger than that computed from astrometry alone,
the added radial velocity information likely supports the belief that the two
stars are truly comoving in 3-space.
Conversely, if the new likelihood ratio is smaller, the radial velocity data
indicate that the pair is likely a false-positive and are not truly comoving.
In practice, we compute the logarithm of the likelihood ratios,
$\mathcal{R} = \ln \mathcal{L}_1/\mathcal{L}_2$; we refer to the
proper-motion-only log-FML-ratio for a given pair of stars as
\llrold\, and the log-FML-ratio including the radial velocity data as
\llrnew.
We consider a pair as being genuinely comoving if $\llrnew > \llrold$; 125 of
the 312 comoving pairs satisfy this constraint.
\figurename~\ref{fig:llr}

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width=\linewidth]{llr-vs-dv.pdf}
  \end{center}
  \caption{%
    FML
    \label{fig:llr}}
\end{figure}

TODO: check on ipython notebook re-running with updated observation shite

@@

\todo{Plot of magnitude of 3d velocity difference vs. 3d separation?}

\todo{separation distribution - compare with Shaya and Olling sample!}
\todo{10 overlap with Shaya and Olling}
